---
title: "Salmonella mouse analysis"
output:
html_notebook: default
---

Load libraries
```{r}
library(dplyr)
library(readr)
library(FluxBalanceAnalyzeR)
library(stringr)
library(tidyr)
library(purrr)
library(ggplot2)
library(magrittr)
library(forcats)
```

Read in reaction table
```{r}
rxns <- read_tsv('../data/LT2_react.tsv') %>%
  select(abbreviation=`Rxn name`,
         equation=Formula,
         uppbnd=UB,
         lowbnd=LB,
         geneAssociation=`Gene-reaction association`,
         Subsystem,
         name = `Rxn description`
  ) %>%
  mutate(obj_coef=1*(abbreviation=='biomass_iRR1083')) %>%
  mutate(geneAssociation = str_replace_all(geneAssociation, 'and', '&'),
         geneAssociation = str_replace_all(geneAssociation, 'or', '|')) %>%
  sample_frac()

```

Custom reaction bounds
```{r}
rxns <- rxns %>%
  mutate(lowbnd = ifelse(abbreviation=='EX_skm(e)',-10,lowbnd))
```

From reaction table, list all genes
```{r}
genes_in_model <- rxns$geneAssociation %>%
  str_split('[()|& ]+') %>%
  flatten_chr() %>%
  discard(is.na) %>%
  discard(~ str_length(.x)==0)
```

Map between STM style codes and gene names
```{r}
genes_mapping <- read_tsv('../data/LT2_genes.tsv')

gene_names_in_model <- genes_mapping %>%
  filter(Gene %in% genes_in_model) %>%
  getElement('Alias')
```

Select which group to use as the control. This will produce a wild type output.
```{r}
control <- 'Input'
```

Read in rna counts and preprocess them
```{r}
rnas <- inner_join(
  read_tsv('../data/seq_counts.tsv') %>%
    rename(Alias = Gene),
  genes_mapping %>%
    select(Alias,
           Gene)
) %>%
  filter(Gene %in% genes_in_model) %>%
  gather(measurement, count, -Locus, -Gene, -Product, -Alias) %>%
  separate(measurement, c('group', 'replicate'), '_') %>%
  mutate(replicate = str_c(group, '_', replicate)) %>%
  # next, remove experimental variance between replicates
  group_by(replicate) %>%
  mutate(count = log(count+1),
         count = count/mean(count)) %>%
  ungroup %>%
  # now, average and normalize groups against control
  group_by(Gene, group) %>%
  summarise(meancount = mean(count)) %>%
  spread(group, meancount) %>%
  ungroup %>%
  mutate(Group1 = Group1/.[[control]],
         Group2 = Group2/.[[control]],
         Group3 = Group3/.[[control]],
         Input = Input/.[[control]]) %>%
  gather('group', 'presence', -Gene) %>% 
  mutate(presence = qnorm(pnorm(presence, mean=1, sd=sd(presence)), mean=1, sd=0.2)) %T>%
  (function(x){print(x %>%
                       ggplot(aes(x=presence, y=qnorm(rank(presence)/length(presence)))) +
                       geom_point()
  )})
```

Project the gene expressions onto the fluxes
```{r}
models_with_expression <- rnas %>%
  group_by(group) %>%
  do(bind_rows(., data_frame(Gene=setdiff(genes_in_model, .$Gene), presence=NA, group=.$group[1]))) %>%
  nest() %>%
  mutate(data = map(data, function(x){
    rxns %>%
      mutate(activation = FluxBalanceAnalyzeR::gene_eval(geneAssociation, x$Gene, x$presence))
  })) %>%
  unnest()
```

adjust flux bounds in line with activation
```{r}
models_with_targets <- models_with_expression %>%
  group_by(group) %>%
  nest %>%
  mutate(data = map(data, find_flux_variability_df)) %>%
  unnest() %>%
  rename(baseflux = flux) %>%
  group_by(abbreviation) %>%
  mutate(
    stable = max(sd)<0.000001 & sd(baseflux) < 0.1*mean(abs(baseflux)),
    activation = ifelse(stable, activation, NA)
  ) %>%
  select(-sd) %>%
  #mutate(activation=1) %>% # for testing
  mutate(
    adjustment = ifelse(is.na(activation), abs(baseflux)*0.1+1, abs(activation-1)*abs(baseflux)*0.1+1),
    uppbnd = ifelse(uppbnd>0, baseflux + adjustment, pmin(0, baseflux + adjustment)),
    lowbnd = ifelse(lowbnd<0, baseflux - adjustment, pmax(0, baseflux - adjustment)),
    adjustment=NULL,
    lowbnd = ifelse(abbreviation=='biomass_iRR1083', baseflux, lowbnd),
    obj_coef = ifelse(is.finite(activation), (activation-1)*sign(baseflux), 0)
    #obj_coef = obj_coef + (abbreviation=='biomass_iRR1083')*1
  ) %>%
  ungroup %T>%
  (function(x){print(x %>%
                       filter(!(is.na(activation))) %>%
                       #filter(abs(baseflux)<900) %>%
                       ggplot(aes(x = baseflux)) + 
                       geom_ribbon(aes(ymin=lowbnd, ymax=uppbnd)) + 
                       geom_point(aes(y=baseflux), colour='blue') + scale_y_continuous(trans = scales::trans_new('cbrt', function(x){sign(x)*abs(x)^(1/3)}, function(x){x^3}))
  )})
```

calculate final fluxes
```{r}
results <- models_with_targets %>%
  group_by(group) %>%
  nest() %>%
  mutate(data = map(data, find_flux_variability_df, do_minimization=FALSE)) %>%
  unnest() %T>%
  (. %>%
     filter(str_detect(abbreviation, 'iomass')) %>%
     select(group, abbreviation, flux) %>% 
     print
  )
```

Produce final figures
=====================

Prepare flux data
```{r}
results_analysed_3 <- results %>%
  # filter out uninteresting
  group_by(abbreviation) %>%
  mutate(
    baseflux = flux[group=='Input'],
    boring = all(near(flux, baseflux)),
    normflux = (flux-baseflux)/(uppbnd-lowbnd),
    stable_1 = max(sd)<0.000001,
    stable_2 = sd(baseflux)<0.001*mean(abs(baseflux))
    ) %>%
  ungroup
```

Plot an overall horizontal summary
```{r}
horizontal_point_summary <- results_analysed_3 %>%
    filter(stable_1, stable_2, !boring) %>%
    filter(Subsystem!='Exchange') %>%
    arrange(Subsystem,abbreviation, group) %>%
    mutate(abbreviation=fct_inorder(abbreviation)) %>%
    # plot
    ggplot(aes(x=normflux, y=abbreviation)) + 
    scale_x_continuous(labels=scales::percent, breaks=seq(-1,1,0.5), minor_breaks=seq(-1,1,0.25), limits=c(-4,1), trans = scales::trans_new('cbrt', function(x){sign(x)*abs(x)^(1/3)}, function(x){x^3})) +
    geom_point(aes(colour=group, shape=group), alpha=0.5) +
    labs(x='Normalized flux change', y='Reaction abbreviation') +
    theme_bw() + 
    geom_tile(aes(width=0.1, x=-0.9, y=abbreviation, fill=letters[as.numeric(as.factor(Subsystem)) %% 8+1]), show.legend = FALSE, alpha=0.2) +
    geom_text(aes(x=-0.9, label = pathway, group=Subsystem), check_overlap=TRUE, data=(. %>% group_by(Subsystem) %>% mutate(pathway = ifelse(abbreviation==nth(abbreviation,round(n()/2)),Subsystem,NA))), hjust=1, alpha=0.5, size=2.5) +
    scale_fill_brewer(type='qual', palette = 'Set3')

ggsave('../publication_images/horizontal_point_summary.png',plot=horizontal_point_summary)
print(horizontal_point_summary)
```

Prepare data for heatmap by adding metabolite data
```{r}
rate_tables <- results_analysed_3 %>%
  filter(stable_1, stable_2, !boring) %>%
  group_by(group) %>%
  by_slice(function(x){
    inner_join(x, expand_reactions(x)) %>%
      mutate(rate = flux * stoich / abs(stoich))
  }) %>%
  with(set_names(.out, group))
```

Plot heatmaps
```{r}
expand.grid(x=c('Group1','Group2','Group3'), y=c('Group1','Group2','Group3','Input'), stringsAsFactors = FALSE) %>%
  filter(x!=y) %>%
  invoke_rows(.d=., .f=function(x,y){
    (inner_join(rate_tables[[x]], rate_tables[[y]], by=c('abbreviation', 'met')) %>%
       mutate(diff = rate.x-rate.y) %>%
       group_by(abbreviation) %>%
       filter(max(abs(diff))>0.05) %>%
       ggplot(aes(x=abbreviation, y=met, fill=diff)) + 
       geom_tile() +
       scale_fill_gradient2(midpoint=0, low='blue', high='red') + 
       labs(x='Reaction', y='Metabolite', fill='Raw flux difference') +
       ggtitle(str_c(x, ' vs ', y)) +
       theme_bw() + 
       theme(axis.text.x = element_text(angle = 45, hjust = 1))
    ) %>%
      ggsave(paste0('../publication_images/heatmap_',x,'_vs_',y,'.png'),.)
  }
  )
```

Output table
```{r}
results_analysed_3 %>%
  filter(stable_1, stable_2, !boring) %>%
  select(abbreviation, group, flux, name, equation, Subsystem, geneAssociation) %>%
  spread(group, flux) %>%
  select(abbreviation, Input, starts_with('Group'), everything()) %>%
  write_tsv('../temp/important.tsv')
```



